\chapter{Transformer Architecture -- The Foundation}
\minitoc

\begin{chapteroverview}
  \begin{itemize}
    \item Evaluate the decoder-only transformer and its 2026 architectural variations.
    \item Analyze MQA, GQA, and MLA for efficient KV cache management at scale.
    \item Compare modern positional encodings including RoPE, iRoPE, and ALiBi.
    \item Review production-standard normalization (RMSNorm) and activation (SwiGLU) functions.
    \item Understand the 2025 transition to fine-grained Mixture of Experts (MoE).
  \end{itemize}
\end{chapteroverview}

\section{The Decoder-Only Transformer}

Nearly all modern autoregressive LLMs use a decoder-only transformer architecture: sequential blocks of masked self-attention and feed-forward sub-layers. In PaLM-540B, approximately 90\% of parameters reside in feed-forward layers---which is why MoE focuses on FFN efficiency.

\subsection{Architectural Comparison: Key Modern Models}

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{2.8cm}L{2cm}L{2cm}L{3.5cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Model}} & \textcolor{white}{\textbf{Params}} & \textcolor{white}{\textbf{Context}} & \textcolor{white}{\textbf{Notable Features}} \\
\midrule
Llama 3.1 (70B) & 70B & 128K & GQA, RoPE, SwiGLU \\
Mistral 7B & 7B & 32K & GQA, sliding window \\
Mixtral 8x7B & 47B total & 32K & MoE, 13B active \\
DeepSeek-V3 & 671B total & 128K & Fine-grained MoE, MLA \\
Llama 4 Maverick & 400B total & 1M & MoE, iRoPE, 128 experts \\
\bottomrule
\end{tabular}
\caption{Representative modern model architectures}
\end{table}

\section{Key Architectural Components}

\subsection{Multi-Head Attention Variants}

Standard multi-head attention (MHA) uses $H$ heads each with independent $Q$, $K$, $V$ projections (\hyperref[form:sdpa]{formulation}: Appendix~\ref{app:attention}; \hyperref[lst:sdpa]{code}: Appendix~H). The KV cache grows as $2 \times H \times d_\text{head} \times L$ per token at inference---a critical bottleneck at long context.

\begin{itemize}
  \item \textbf{Multi-Query Attention (MQA):} Single K,V head shared by all query heads. Maximum KV cache reduction; can hurt quality at small scale (\hyperref[form:mqa]{formulation \S\ref*{form:mqa}}).
  \item \textbf{Grouped Query Attention (GQA):} $G$ KV head groups ($1 < G < H$) \parencite{ainslie2023gqa}. Balances cache reduction and quality. Used in Llama 3, Mistral, and most 2025 models (\hyperref[form:gqa]{formulation \S\ref*{form:gqa}}, \hyperref[lst:gqa]{code Listing~\ref*{lst:gqa}}).
  \item \textbf{Multi-head Latent Attention (MLA):} DeepSeek-V3. KV cache is compressed into a low-rank latent vector $c_{KV}$, from which heads are projected via an up-projection matrix $W_{UK}$. This achieves the memory footprint of MQA while maintaining the expressive power of MHA (see \hyperref[form:mla]{formulation \S\ref*{form:mla}}, \hyperref[lst:mla]{code Listing~\ref*{lst:mla}}).
\end{itemize}

\begin{keyinsight}[MLA Mechanics]
MLA's core innovation is the decoupled latent vector:
\[
c_{KV} = W_{DK} h_t, \quad k_t, v_t = W_{UK} c_{KV}
\]
where $W_{DK}$ is a down-projection and $W_{UK}$ an up-projection. RoPE is applied to a separate, non-compressed portion of the query and key to maintain relative position awareness without bloating the KV cache.
\end{keyinsight}

\subsection{Positional Encodings}

Full derivations in Appendix~\ref{app:pos}.
\begin{itemize}
  \item \textbf{RoPE (Rotary Position Embedding):} Applies rotation matrices to Q/K vectors \parencite{su2024roformer}. Encodes relative positions implicitly. Dominant 2025--2026 standard.
  \item \textbf{iRoPE (Llama 4):} Interleaved no-position and RoPE layers. Enables 10M-token context without explicit positional interpolation.
  \item \textbf{ALiBi:} Linear bias on attention scores \parencite{press2022train}. Zero extra parameters, strong length generalization.
  \item \textbf{YaRN / LongRoPE:} NTK-aware RoPE interpolation for 4--8$\times$ context extension without full retraining \parencite{peng2023yarn}.
\end{itemize}

\subsection{Normalization and Activation}

Full formulations in Appendix~\ref{app:norm} (\hyperref[form:rmsnorm]{RMSNorm}, \hyperref[form:swiglu]{SwiGLU}).
\begin{itemize}
  \item \textbf{RMSNorm with pre-normalization:} More stable than post-norm LayerNorm. Dominant in all major 2025 models (\hyperref[form:rmsnorm]{formulation \S\ref*{form:rmsnorm}}).
  \item \textbf{SwiGLU:} Gated FFN activation combining Swish and GLU \parencite{shazeer2020glu}. Standard for 2024--2026 (\hyperref[form:swiglu]{formulation \S\ref*{form:swiglu}}).
\end{itemize}

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{2cm}L{3.5cm}L{5cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Function}} & \textcolor{white}{\textbf{Formulation}} & \textcolor{white}{\textbf{Key Trade-off / Usage}} \\
\midrule
ReLU & $\max(0, x)$ & Simple, fast. Risk of ``dying ReLU'' (zero gradient). \\
GeLU & $x \Phi(x)$ & Smooth, probabilistic gating. GPT-2/3 standard. \\
Swish / SiLU & $x \sigma(x)$ & Non-monotonic, smoother than ReLU. Llama 1/2. \\
SwiGLU & $\text{Gated}(x, \text{Swish})$ & Most expressive. 2026 production standard. \\
\bottomrule
\end{tabular}
\caption{Comparative analysis of activation functions in LLMs}
\end{table}

\section{Mixture of Experts (MoE)}

Replace the dense FFN with multiple smaller expert networks and a router selecting which experts process each token \parencite{shazeer2017outrageously} (\hyperref[form:moe]{formulation \S\ref*{form:moe}}, \hyperref[lst:moe]{code Listing~\ref*{lst:moe}}). Only 1--2 of 8--64 experts activate per token. Mixtral 8x7B: 47B total parameters, $\sim$13B active per forward pass.

\begin{keyinsight}[Key MoE Developments in 2025--2026]
\begin{itemize}
  \item \textbf{DeepSeekMoE:} Fine-grained experts with shared expert isolation. 256 routed experts + 1 shared expert per layer.
  \item \textbf{DeepSeekMoE \parencite{dai2024deepseekmoe}:} Fine-grained experts with shared expert isolation. 256 routed experts + 1 shared expert per layer.
  \item \textbf{Sparse Upcycling:} Convert dense models to MoE without training from scratch.
  \item \textbf{SwitchHead:} MoE applied to attention projection layers (Q, K, V).
  \item \textbf{Expert Parallelism:} Distributed experts across GPUs for efficient scale.
  \item \textbf{Load Balancing:} Auxiliary loss or expert-choice routing prevents expert collapse.
\end{itemize}
\end{keyinsight}

\section{Context Length and Efficient Attention}

\begin{itemize}
  \item \textbf{Flash Attention (v2/v3):} $O(n^2) \rightarrow O(n)$ memory via IO-aware tiling \parencite{dao2022flashattention,dao2023flashattention2,shah2024flashattention3}. Mandatory for modern training. FA3 adds warp-specialization for $\sim$2$\times$ speedup over FA2 on H100.
  \item \textbf{Ring Attention:} Million-token contexts across devices in a ring topology. Each device holds one segment; KV chunks circulate.
  \item \textbf{Sliding Window Attention:} Fixed local window with cross-layer information flow (Mistral).
\end{itemize}

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{3.5cm}C{2.5cm}L{4.5cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Context Window}} & \textcolor{white}{\textbf{KV Cache (7B)}} & \textcolor{white}{\textbf{Use Case}} \\
\midrule
4K tokens & 512 MB & Short conversations \\
32K tokens & 4 GB & Document processing \\
128K tokens & 16 GB & Full codebase analysis \\
1M tokens & 128 GB & Entire book corpus \\
\bottomrule
\end{tabular}
\caption{Approximate KV cache memory at different context lengths (BF16, 32 heads)}
\end{table}


% ══════════════════════════════════════════════════════════════════
%  PART II: DATA
% ══════════════════════════════════════════════════════════════════
