\part{Data -- The Foundation of Everything}

\chapter{Data Curation -- The Oil for the Engine}
\minitoc

\begin{chapteroverview}
  \begin{itemize}
    \item Implement quality filtering and deduplication at the trillion-token scale.
    \item Review domain sampling proportions (Web, Code, Science, Books) for 2026.
    \item Analyze strategies for removing PII and toxic content while preserving knowledge.
    \item Understand the 2025 shift from quantity-driven to quality-driven curation.
  \end{itemize}
\end{chapteroverview}

\section{Why Data Quality Trumps Quantity}

FineWeb \parencite{penedo2024fineweb} (15T tokens, heavily curated) outperforms RedPajama (20T tokens, lightly filtered)---5 trillion extra noisy tokens \emph{hurt} performance. Modern frontier models train on 10--20T tokens, approaching available high-quality text limits.

\section{Data Sources}

Web crawl (FineWeb, DCLM), books and academic literature, code repositories (The Stack), curated knowledge bases (Wikipedia), and multilingual sources (FineWeb-2).

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{3cm}L{2.5cm}L{5cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Source}} & \textcolor{white}{\textbf{Typical Mix}} & \textcolor{white}{\textbf{Notes}} \\
\midrule
Web crawl & 50--70\% & Requires aggressive cleaning \\
Code & 15--25\% & Boosts reasoning and instruction following \\
Books & 5--15\% & Long-form coherence \\
Scientific papers & 3--8\% & Technical reasoning \\
Wikipedia / Wikidata & 2--5\% & Factual grounding \\
Curated instruction data & 1--3\% & Bridges to SFT \\
\bottomrule
\end{tabular}
\caption{Typical data source composition for pre-training}
\end{table}

\section{Cleaning and Filtering Pipeline}

\subsection{Levels 0--3: Rule-Based}
URL filtering, language identification, boilerplate removal, heuristic quality filters (perplexity thresholds, line length statistics, repetition detection).

\subsection{Levels 4--5: Model-Based}
FastText/transformer classifiers for quality scoring. GneissWeb: sharded deduplication at 10T scale. SoftDedup: reweight rather than delete---preserves rare but informative examples.

\subsection{Levels 6+: LLM-Driven (2025--2026 Frontier)}
DataEvolve / Data Darwinism: autonomous LLM-driven quality analysis and iterative cleaning. Category-specific strategies (SwallowCode, MegaMath). Instruction-response augmented pre-training.

\section{Deduplication}

Exact (hash-based), near-duplicate (MinHash/LSH with GPU-accelerated connected components), and fuzzy (embedding-based similarity). Near-deduplication at the paragraph level catches reformatted reposts that exact dedup misses.

\section{Multi-Stage Training}

Stage 1: large corpus (10T+) for breadth. Stage 2: curated subset (0.5--1T) for depth and polish. Some labs add a Stage 3: task-specific mix (code-heavy, math-heavy) aligned to deployment use case.

\begin{warningbox}[Data Contamination]
Evaluation benchmarks present in pre-training data inflate benchmark scores without improving actual capability. Decontaminate training data by removing near-exact matches to test sets. Always report decontamination procedures in training reports.
\end{warningbox}
