\chapter{Synthetic Data Generation}
\minitoc

\begin{chapteroverview}
  \begin{itemize}
    \item Review the self-instruct and prompt-based data generation evolution.
    \item Master Magpie and PersonaHub for generating high-diversity instruction sets.
    \item Implement quality control and model collapse prevention strategies.
    \item Analyze the cost-efficiency of synthetic data vs. human annotation.
  \end{itemize}
\end{chapteroverview}

Frontier models are trained on trillions of tokens---but high-quality human-written text is finite. Synthetic data, generated or augmented by existing LLMs, now supplies a substantial fraction of both pre-training and post-training corpora.

\section{The Case for Synthetic Data}

\begin{itemize}
  \item \textbf{Scale without annotation cost:} A single human expert can label $\sim$1K examples per day; a model can generate millions in minutes.
  \item \textbf{Long-tail coverage:} Rare instructions, specialized domains, and unusual reasoning patterns are under-represented in web crawls.
  \item \textbf{Consistency:} Synthetic data can be templated to enforce format, difficulty, and safety constraints uniformly.
  \item \textbf{Reasoning chain density:} Writing out step-by-step solutions is prohibitively slow for humans; LLMs produce them at scale.
\end{itemize}

\section{Core Techniques}

\subsection{Self-Instruct and Alpaca}

Self-Instruct \parencite{wang2022selfinstructaligning} bootstraps instruction data from a seed set of 175 human-written examples, using an LLM to generate new (instruction, input, output) triples, then filtering by ROUGE-based deduplication. Stanford Alpaca \parencite{taori2023alpaca} applied this to LLaMA-7B using GPT-3.5, generating 52K instruction-following examples for under \$500. Despite the low cost, Alpaca matched InstructGPT-175B on several instruction-following metrics, demonstrating the asymmetric leverage of synthetic data.

\subsection{Evol-Instruct (WizardLM)}

Rather than generating new instructions from scratch, Evol-Instruct \parencite{xu2023wizardlm} mutates an existing instruction through four ``evolution'' operators: add constraints, deepen complexity, concretize, increase reasoning steps. Each mutation produces a harder variant. Training on evolved instructions produces models that significantly outperform models trained on the original dataset because they see a graduated difficulty curriculum automatically.

\subsection{Magpie: Model Self-Generation}

Magpie \parencite{xu2024magpie} exploits the observation that instruction-tuned models exposed to only a system prompt and role prefix will auto-complete realistic user instructions. The pipeline:
\begin{enumerate}
  \item Feed the model only the system header + user-role prefix (e.g., \texttt{<|start\_header\_id|>user<|end\_header\_id|>}).
  \item Sample a user instruction from the model's own distribution.
  \item Feed the (system + generated instruction) back, sample the assistant response.
  \item Filter by quality score (self-consistency, LLM-judge).
\end{enumerate}
Magpie produces on-distribution synthetic data without any seed examples and scales to millions of examples cheaply. Models fine-tuned on Magpie data at the 8B scale have ranked competitively on AlpacaEval 2.

\subsection{Orca-Style Chain-of-Thought Augmentation}

Orca \parencite{mukherjee2023orca} augments a dataset of short instruction-response pairs from FLAN by adding \emph{detailed reasoning traces} generated by GPT-4. The key insight: small models benefit more from observing the reasoning process than from observing the final answer alone. Orca-2 and later OpenHermes models confirm that 200K--1M GPT-4-augmented examples with reasoning traces can match much larger instruction-tuned models.

\subsection{Synthetic Math and Code Chains}

For verifiable domains, synthetic chains can be validated automatically:
\begin{itemize}
  \item \textbf{Math (MATH, GSM8K-style):} Generate problems via template + parameter randomization. Solve with a CAS (SymPy, Wolfram) to verify. Filter chains where the LLM-generated solution matches the verified answer.
  \item \textbf{Code:} Generate function docstrings and unit tests first, then generate implementations. Execute tests to verify correctness. Only passing solutions enter training data.
\end{itemize}
DeepSeek's reasoning data pipeline uses this approach: generate a candidate, verify via code execution or symbolic math, keep correct examples with their reasoning chains.

\subsection{Persona-Driven Diversity}

Persona Hub \parencite{chan2024personahub}: maintain a library of 1 billion synthesized user personas (age, profession, goals, style). Sample personas at generation time to produce diverse instruction styles. Prevents the ``assistant voice'' homogenization that afflicts single-persona synthesis.

\section{Quality Filtering}

Raw synthetic data contains errors, repetition, and unsafe content. A multi-stage filter is essential:

\begin{enumerate}
  \item \textbf{Format checks:} Regex validation for expected chat template structure.
  \item \textbf{Length filtering:} Remove excessively short responses ($<$20 tokens) and runaway generations ($>$4K tokens without meaningful content).
  \item \textbf{Self-consistency scoring:} Generate the response twice with different seeds; low cosine similarity flags inconsistent or hallucinated content.
  \item \textbf{LLM-as-judge:} Score on helpfulness, accuracy, and safety (1--5 scale). Discard scores $\leq$3.
  \item \textbf{Deduplication:} MinHash near-deduplication to prevent overfitting to repeated patterns.
  \item \textbf{Human spot-check:} 1--2\% random sample reviewed by domain experts to calibrate automated filters.
\end{enumerate}

\section{Risks: Model Collapse and Bias Amplification}

\begin{warningbox}[Model Collapse]
Training repeatedly on model-generated data without human-curated anchors leads to \emph{model collapse} \parencite{shumailov2024ai}: the distribution narrows, variance decreases, and tail behaviors (rare but important knowledge) are lost. The antidote: always maintain a human-curated anchor set---typically 10--30\% of the training mix---and monitor perplexity on a fixed held-out real-text corpus. If perplexity rises on the real-text held-out set, collapse is beginning.
\end{warningbox}

Bias amplification: errors and biases in the teacher model are reproduced and reinforced in the student. Use diverse teacher models (not just one API) and human review to audit for systematic failures.

\section{Synthetic Data Composition Recommendations}

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{3.5cm}L{2.5cm}L{5cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Synthetic Source}} & \textcolor{white}{\textbf{Mix \%}} & \textcolor{white}{\textbf{Purpose}} \\
\midrule
Magpie (self-generated) & 20--40\% & On-distribution instruction diversity \\
Evol-Instruct evolved & 15--25\% & Difficulty calibration \\
Verified math/code chains & 10--20\% & Reasoning capability \\
Orca CoT augmented & 10--20\% & Step-by-step process learning \\
Human-curated anchor & 20--30\% & Distribution grounding, collapse prevention \\
\bottomrule
\end{tabular}
\caption{Recommended SFT data mix composition using synthetic sources}
\end{table}


% ══════════════════════════════════════════════════════════════════
%  PART III: PRE-TRAINING
% ══════════════════════════════════════════════════════════════════
