\chapter{Alignment -- RLHF, DPO, and Beyond}
\minitoc

\begin{chapteroverview}
  \begin{itemize}
    \item Trace the evolution from RLHF to reference-free methods (DPO, SimPO).
    \item Implement alignment with binary feedback (KTO) and group-relative rewards (GRPO).
    \item Evaluate the role of verifiable rewards (RLVR) in reasoning-heavy models.
    \item Understand the 2026 modular alignment stack for safety and helpfulness.
  \end{itemize}
\end{chapteroverview}

\section{The Alignment Problem}

A grammatically perfect model may be unhelpful, evasive, or unsafe. The field has evolved from RLHF alone to a modular stack of complementary methods.

\section{Reward Modeling}

Complete objectives in Appendix~\ref{app:alignment}: \hyperref[form:ppo]{PPO}, \hyperref[form:dpo]{DPO}, \hyperref[form:simpo]{SimPO}, \hyperref[form:kto]{KTO}, \hyperref[form:grpo]{GRPO}.

The reward model (RM) translates human preferences into scalar scores. RM quality determines the ceiling for all RM-based alignment methods.

\begin{itemize}
  \item \textbf{Architecture:} Language model with a regression head replacing the LM head.
  \item \textbf{Training:} Bradley-Terry model on pairwise comparisons \parencite{ziegler2019finetuning}: $p(y_w \succ y_l) = \sigma(r(x, y_w) - r(x, y_l))$.
  \item \textbf{Reward hacking:} An RM optimized too strongly diverges from true human preferences. KL penalty and early stopping mitigate this.
  \item \textbf{Process Reward Models (PRM) \parencite{lightman2023lets}:} Score each intermediate reasoning step, not just the final output. More signal, less sparse. Used in OpenAI's o-series for math and coding.
\end{itemize}

\section{RLHF}

Transformed GPT-3 into ChatGPT \parencite{ouyang2022training}. Pipeline: SFT $\rightarrow$ Reward Model $\rightarrow$ PPO \parencite{schulman2017proximal}. Essential for open-ended quality but requires four models in memory simultaneously and is prone to reward hacking.

\section{DPO}

Eliminates the reward model entirely \parencite{rafailov2023direct} (\hyperref[form:dpo]{formula \S\ref*{form:dpo}}, \hyperref[lst:dpo]{code Listing~\ref*{lst:dpo}}). 90--95\% of RLHF performance at 40--60\% less compute.

\begin{keyinsight}[DPO Successors]
\begin{itemize}
  \item \textbf{SimPO \parencite{meng2024simpo}:} Eliminates the reference model entirely. It uses length-normalized log probabilities as an implicit reward and introduces a target reward margin $\gamma$ (see \hyperref[form:simpo]{formulation \S\ref*{form:simpo}}, \hyperref[lst:simpo]{code Listing~\ref*{lst:simpo}}).
  \item \textbf{KTO (Kahneman-Tversky Optimization) \parencite{ethayarajh2024kto}:} Operates on binary thumbs-up/down signals rather than preferences. It maximizes the utility of model outputs by aligning with human loss aversion (see \hyperref[form:kto]{formulation \S\ref*{form:kto}}, \hyperref[lst:kto]{code Listing~\ref*{lst:kto}}).
  \item \textbf{Iterative DPO:} Semi-online re-sampling from the current policy. Closes the distribution gap between training data and current model outputs.
  \item \textbf{IPO \parencite{azar2024ipo}:} Information-Preference Optimization. Regularizes DPO to prevent overfitting to preference margins.
\end{itemize}
\end{keyinsight}

\section{GRPO}

DeepSeek's dominant algorithm for reasoning \parencite{guo2025deepseekr1} (\hyperref[form:grpo]{formula \S\ref*{form:grpo}}, \hyperref[lst:grpo]{code Listing~\ref*{lst:grpo}}). Generates a group of 8--64 responses per prompt. Advantages normalized against group statistics. Eliminates the critic model, yielding 33--50\% memory savings over PPO.

\[
\mathcal{L}_\text{GRPO} = \mathbb{E}\left[\sum_i \hat{A}_i \cdot \log \pi_\theta(o_i \mid q) - \beta \cdot \text{KL}[\pi_\theta \| \pi_\text{ref}]\right]
\]

where $\hat{A}_i = \frac{r_i - \text{mean}(r)}{\text{std}(r)}$ is the group-normalized advantage.

\section{DAPO}

Removes reference model dependency from GRPO \parencite{yu2025dapo}. The most streamlined RL alignment approach. Also introduces \emph{clip-higher} (asymmetric policy clipping) and token-level loss normalization to stabilize training.

\section{RLAIF and Constitutional AI}

AI-generated preferences replace human annotations. Anthropic's CAI \parencite{bai2022constitutional}: model self-critiques against constitutional principles, creating scalable alignment data without per-example human annotation.

\section{RLVR}

Verifiable rewards for math and code. Binary, objective signal resists reward hacking. DeepSeek R1 demonstrated emergent reasoning from pure RLVR + GRPO without SFT. The key insight: verifiable correctness is a cleaner training signal than a learned reward model.

\section{The Modern Alignment Stack}

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{2.8cm}L{3.5cm}L{4.2cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Stage}} & \textcolor{white}{\textbf{Technique}} & \textcolor{white}{\textbf{What It Solves}} \\
\midrule
Instruction Following & SFT & Format, tone, task execution \\
Preference Alignment & DPO / SimPO / KTO & Human values, norms \\
Reasoning & GRPO / DAPO + RLVR & Math, code, planning \\
Safety \& Helpfulness & RLHF (neural RM) & Open-ended quality, harm avoidance \\
Constitutional Guardrails & CAI / RLAIF & Scalable safety principles \\
\bottomrule
\end{tabular}
\caption{The modular alignment stack in 2026}
\end{table}
