\chapter{Knowledge Distillation}

Knowledge distillation (KD) transfers capability from a large \emph{teacher} model to a smaller \emph{student}, achieving performance beyond what the student could reach by training on ground-truth labels alone.

\section{Why Distillation Matters}

Frontier labs use distillation systematically: Behemoth $\rightarrow$ Scout/Maverick (Meta), R1-671B $\rightarrow$ 1.5B--70B (DeepSeek), Gemini Pro $\rightarrow$ Flash $\rightarrow$ Nano (Google). Distilled 7B models now routinely outperform undistilled 70B models on target tasks.

\section{Response (Output-Level) Distillation}

The simplest form: generate responses from the teacher and use them as SFT labels for the student. The student minimizes the negative log-likelihood of teacher completions rather than human-annotated completions.

\[
\mathcal{L}_\text{resp} = -\sum_t \log p_\theta(y_t^\text{teacher} \mid x, y_{<t}^\text{teacher})
\]

\textbf{Advantages:} Simple pipeline; works with closed-source teachers (you only need teacher outputs, not logits). \textbf{Disadvantage:} Ignores the full teacher probability distribution---the student does not learn ``how uncertain'' the teacher is.

\section{Logit-Level Distillation}

When teacher logits are available (open-weight teachers), the student matches the teacher's full output distribution via KL divergence:

\[
\mathcal{L}_\text{KD} = \alpha \cdot \mathcal{L}_\text{CE}(y^\text{gt}, p_\theta) + (1-\alpha) \cdot T^2 \cdot D_\text{KL}\!\left(\text{softmax}(z^\text{teacher}/T) \;\|\; \text{softmax}(z^\text{student}/T)\right)
\]

where $T$ is the temperature (commonly 2--4) and $\alpha$ balances ground-truth and teacher loss. Higher temperature softens the teacher distribution, providing richer signal on near-correct tokens.

\section{Sequence-Level Knowledge Distillation (SeqKD)}

Rather than optimizing token-level KL, SeqKD \parencite{kim2016sequence} generates \emph{mode sequences} from the teacher (beam search, temperature $= 0$) and treats them as hard labels. More cache-friendly than online logit distillation; works without access to logits. Used extensively when distilling from GPT-4 or Claude.

\section{MiniLLM: Reverse KL for Distillation}

Standard KD minimizes forward KL $D_\text{KL}(p_\text{teacher} \| p_\text{student})$, which forces the student to cover all teacher modes and leads to ``mean-seeking'' behaviour. MiniLLM \parencite{gu2024minillm} minimizes reverse KL $D_\text{KL}(p_\text{student} \| p_\text{teacher})$, encouraging mode-seeking: the student becomes sharp and accurate on the teacher's most probable outputs, avoiding the diffusion of probability mass across irrelevant alternatives. Reverse KL is estimated via policy gradient on student-generated sequences with teacher log-probability as reward.

\section{Intermediate Layer Distillation}

Align hidden states or attention patterns at intermediate layers, not just the output:

\begin{itemize}
  \item \textbf{Feature-map distillation (PKD, TinyBERT):} $\mathcal{L}_\text{feat} = \|W_\text{proj} h^\text{student}_l - h^\text{teacher}_{f(l)}\|_2^2$ where $f(l)$ maps student layer to teacher layer index and $W_\text{proj}$ is a learned projection.
  \item \textbf{Attention-pattern distillation:} $\mathcal{L}_\text{attn} = D_\text{KL}(A^\text{teacher}_{l,h} \| A^\text{student}_{l,h})$ summed over heads and layers.
  \item \textbf{Practical note:} Intermediate distillation requires matching architectures or learned projections. Most practical pipelines stick to output-level or logit-level distillation.
\end{itemize}

\section{On-Policy Distillation}

A key failure mode of offline response distillation: the student is trained on teacher trajectories but evaluated on \emph{its own} trajectories. Distribution mismatch grows with model capability gap. On-policy distillation:

\begin{enumerate}
  \item Roll out the \emph{student} to generate candidate completions.
  \item Score each completion with the teacher (assign log-probability or RM score).
  \item Minimize KL between student and teacher-scored distribution.
\end{enumerate}

Requires either white-box access to teacher logits or a fast teacher inference endpoint. Used in GKD \parencite{agarwal2024onpolicy} (Generalized Knowledge Distillation).

\section{Speculative Decoding as Distillation}

Speculative decoding \parencite{leviathan2023fast} uses a small draft model to propose $k$ tokens, verified in parallel by the target model. Beyond a decoding speedup, training the drafter to maximally predict acceptance by the target is a form of distillation---the drafter learns to approximate the target's conditional distribution on easy tokens, which constitute 70--90\% of all tokens.

\section{Practical Distillation Workflow}

\begin{enumerate}
  \item Select teacher (strongest accessible model: open-weight 70B--671B or API).
  \item Generate responses on your instruction set. For reasoning tasks, generate with high temperature ($T = 0.7$--$1.0$) and filter by verifiable correctness.
  \item Optionally: collect teacher logits for logit-level KD (requires local teacher).
  \item Fine-tune student with $\alpha = 0.5$: 50\% KD loss, 50\% standard CE loss on human labels.
  \item Evaluate on target benchmarks; iterate on temperature and $\alpha$.
\end{enumerate}

\begin{keyinsight}[Distillation Rules of Thumb]
\begin{itemize}
  \item A 7B student distilled from a 70B teacher typically closes 60--80\% of the gap to the teacher.
  \item Distillation data volume: 50K--500K high-quality teacher completions is sufficient for most fine-tuning objectives.
  \item For reasoning tasks, only keep teacher traces that lead to correct verified answers---incorrect reasoning chains hurt more than they help.
  \item On-policy distillation adds significant complexity; prefer offline SeqKD unless you observe severe distribution mismatch.
\end{itemize}
\end{keyinsight}

\begin{lstlisting}[style=pythonstyle, caption={Logit-level knowledge distillation training step}]
import torch
import torch.nn.functional as F

def distillation_loss(student_logits, teacher_logits,
                      labels, alpha=0.5, temperature=2.0):
    """
    Combined cross-entropy + KL distillation loss.
    student_logits, teacher_logits: (batch, seq_len, vocab)
    labels: (batch, seq_len)  -100 for masked tokens
    """
    # Standard cross-entropy on ground-truth labels
    ce_loss = F.cross_entropy(
        student_logits.view(-1, student_logits.size(-1)),
        labels.view(-1),
        ignore_index=-100,
    )

    # KL divergence on teacher distribution
    soft_teacher = F.softmax(teacher_logits / temperature, dim=-1)
    soft_student = F.log_softmax(student_logits / temperature, dim=-1)
    kd_loss = F.kl_div(
        soft_student, soft_teacher,
        reduction="batchmean", log_target=False,
    ) * (temperature ** 2)

    return alpha * ce_loss + (1 - alpha) * kd_loss
\end{lstlisting}


% ══════════════════════════════════════════════════════════════════
%  PART V: EVALUATION & DEPLOYMENT
% ══════════════════════════════════════════════════════════════════
