\part{Evaluation, Safety \& Deployment}

\chapter{Evaluation and Benchmarking}

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{3cm}L{4.5cm}L{2.5cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Benchmark}} & \textcolor{white}{\textbf{What It Measures}} & \textcolor{white}{\textbf{Type}} \\
\midrule
MMLU / MMLU-Pro & Broad academic knowledge (57+ subjects) & Multiple choice \\
HumanEval / MBPP & Code generation correctness & Code execution \\
GSM8K / MATH & Mathematical problem solving & Verified answers \\
TruthfulQA & Resistance to misconceptions & Open-ended \\
MT-Bench & Multi-turn conversation & LLM-judged \\
AlpacaEval 2 & Instruction-following quality & LLM-judged \\
Arena-Hard & Head-to-head comparison & LLM-judged \\
SWE-bench & Real-world software engineering & Code execution \\
AIME 2024 & Competition mathematics & Verified answers \\
GPQA-Diamond & Expert-level science reasoning & Multiple choice \\
\bottomrule
\end{tabular}
\caption{Standard LLM evaluation benchmarks}
\end{table}

\section{LLM-as-Judge}

LLM judges (GPT-4, Claude, Gemini) score open-ended responses. Pitfalls: position bias (judges prefer the first response), verbosity bias (longer responses rated higher), self-enhancement bias (models rate their own outputs higher). Mitigations: swap positions across evaluations, length-normalize scores, use multiple judge models.

\section{Chatbot Arena and ELO Rankings}

Chatbot Arena (LMSYS) collects millions of pairwise human comparisons and computes ELO ratings. Statistically robust due to scale; reflects real user preferences better than academic benchmarks. High ELO and high MMLU are different things---they measure different aspects of model quality.

\section{Benchmark Contamination and Gaming}

\begin{itemize}
  \item Evaluation data present in pre-training inflates scores. Always report decontamination procedures.
  \item Optimizing specifically for a benchmark can improve scores without improving underlying capability (Goodhart's Law).
  \item Mitigations: dynamic benchmarks with new questions each run, held-out test sets never used for selection, diverse multi-benchmark suites that resist narrow optimization.
\end{itemize}
