\chapter{Safety, Ethics, and Constitutional AI}
\minitoc

\begin{chapteroverview}
  \begin{itemize}
    \item Implement red-teaming and adversarial testing for "jailbreak" prevention.
    \item Master Constitutional AI and the RLAIF (RL from AI Feedback) stack.
    \item Analyze mitigation strategies for bias, hallucination, and PII leakage.
    \item Review the 2026 safety standards for high-stakes AI deployments.
  \end{itemize}
\end{chapteroverview}

\section{Safety Taxonomy}

LLM safety failures fall into three broad categories:
\begin{enumerate}
  \item \textbf{Harmfulness:} Generating dangerous, illegal, or offensive content.
  \item \textbf{Dishonesty:} Hallucinating facts, sycophantic agreement, deceptive framing.
  \item \textbf{Misalignment:} Pursuing goals misaligned with user or societal intent.
\end{enumerate}

\section{Constitutional AI (CAI)}

Anthropic's Constitutional AI provides a scalable alternative to extensive human annotation:
\begin{enumerate}
  \item \textbf{SL-CAI:} For each harmful response, the model generates a critique against a written constitution, then revises the response. Revised outputs become SFT training data.
  \item \textbf{RL-CAI:} AI-generated preference labels (RLAIF) replace human annotators for training the reward model. Constitutional principles guide preference generation.
\end{enumerate}

Advantages: dramatically reduces human annotation cost; the constitution is auditable and adjustable; generalizes to unseen harm categories.

\section{Red-Teaming}

\begin{itemize}
  \item \textbf{Manual red-teaming:} Human adversarial testers find novel jailbreaks and failure modes that automated methods miss.
  \item \textbf{Automated red-teaming:} LLM generates adversarial prompts at scale (HarmBench, AdvBench). Identifies systematic weaknesses efficiently.
  \item \textbf{Structured output red-teaming:} Test tool-use, JSON outputs, and code execution paths for injection vulnerabilities.
\end{itemize}

\section{Interpretability and Explainability}

SHAP and LIME for feature attribution. Attention visualization (with caveats---attention patterns are not explanations). Mechanistic interpretability (Anthropic): identify circuits and features responsible for specific behaviors. EU AI Act (2024--2026) mandates transparency for high-risk AI applications.

\section{Safe Deployment Practices}

\begin{itemize}
  \item Output classifiers (toxicity, PII detection) as guardrails.
  \item Input filtering for known harmful patterns.
  \item Rate limiting and usage monitoring.
  \item Human-in-the-loop for high-stakes decisions.
  \item Staged rollouts with safety monitoring before broad deployment.
\end{itemize}
