\chapter{Inference and Deployment at Scale}
\minitoc

\begin{chapteroverview}
  \begin{itemize}
    \item Distinguish between throughput-bound and memory-bound inference.
    \item Master KV cache management (PagedAttention, RadixAttention).
    \item Implement the 2026 quantization stack (H100 FP8, post-training quantization).
    \item Optimize throughput with speculative decoding and drafting.
  \end{itemize}
\end{chapteroverview}

\section{Quantization}

\begin{table}[H]
\centering\small\sffamily
\rowcolors{2}{tablealt}{white}
\begin{tabular}{L{2.5cm}L{2cm}L{3cm}L{3cm}}
\toprule
\rowcolor{tablehead}\textcolor{white}{\textbf{Format}} & \textcolor{white}{\textbf{Size Reduction}} & \textcolor{white}{\textbf{Quality Loss}} & \textcolor{white}{\textbf{Best For}} \\
\midrule
FP16 / BF16 & 2$\times$ & Minimal & Training \\
INT8 (LLM.int8\nolinebreak\parencite{dettmers2022llmint8}) & $\sim$4$\times$ & Negligible & Serving large models \\
INT4 / GPTQ / AWQ & $\sim$8$\times$ & Small & Throughput-optimized serving \\
GGUF (Q4\_K\_M) & $\sim$6$\times$ & Small & Local / edge inference \\
FP8 & 4$\times$ & Minimal & H100/H200 native serving \\
1.58-bit (BitNet) & $\sim$20$\times$ & Moderate & Research / edge \\
\bottomrule
\end{tabular}
\caption{Quantization formats and trade-offs}
\end{table}

\section{Inference Frameworks}

vLLM (PagedAttention), TGI (Hugging Face), TensorRT-LLM (NVIDIA), llama.cpp/Llamafile (local/edge).

\section{Speculative Decoding}

A small draft model generates $k$ tokens; the large target model verifies all $k$ in parallel \parencite{leviathan2023fast}. 2--3$\times$ speedup, zero quality loss. Works best when the draft model acceptance rate is high. Medusa: multiple parallel draft heads on the target model itself, eliminating the separate draft model.

\section{Continuous Batching}

Naive batching waits for all requests in a batch to complete before accepting new ones---wasteful when requests differ in length. Continuous batching \parencite{yu2022orca} inserts new requests as slots free up, achieving near-100\% GPU utilization.

\section{KV Cache Management}

KV cache formula (\hyperref[form:kvcache]{\S\ref*{form:kvcache}}), quantization bounds (\hyperref[form:quant]{\S\ref*{form:quant}}) in Appendix~\ref{app:pipeline}; speculative decoding code: \hyperref[lst:spec_dec]{Listing~\ref*{lst:spec_dec}}.

\begin{itemize}
  \item \textbf{PagedAttention (vLLM) \parencite{kwon2023efficient}:} Non-contiguous physical KV cache blocks managed like OS virtual memory. Eliminates fragmentation, enables 2--4$\times$ more concurrent requests.
  \item \textbf{Prefix caching:} Cache shared system prompt KV states across requests. 50--80\% cache hit rates for common deployments.
  \item \textbf{KV quantization:} Quantize cached KV to INT8/INT4 to extend effective context within fixed memory.
\end{itemize}

\section{Multi-GPU Inference}

For models exceeding single-GPU VRAM, use tensor parallelism at inference:
\begin{itemize}
  \item 2--4 GPUs: tensor parallelism within a node (NVLink bandwidth sufficient).
  \item 8+ GPUs: tensor + pipeline parallelism.
  \item Reference: a 70B model at BF16 requires 4$\times$A100-80GB; INT4 fits on 2$\times$A100-40GB.
\end{itemize}


% ══════════════════════════════════════════════════════════════════
%  PART VI: ADVANCED TOPICS
% ══════════════════════════════════════════════════════════════════
