\part{Advanced Topics and Future Directions}

\chapter{Domain-Specific and Multimodal Models}

\section{Domain-Specific Pre-training}

Domain-specific models outperform general models on target tasks because domain vocabulary and reasoning patterns require exposure during pre-training, not just fine-tuning.

\begin{itemize}
  \item \textbf{BloombergGPT \parencite{wu2023bloomberggpt} (finance):} 50\% financial text, 50\% general. Outperforms GPT-3 on financial benchmarks while remaining competitive on general tasks. Key lesson: 50/50 mix preserves generality.
  \item \textbf{Med-PaLM 2 \parencite{singhal2023large} (medical):} Expert-level performance on the USMLE. Multi-step reasoning over clinical evidence requires both medical knowledge and reasoning capabilities.
  \item \textbf{CodeLlama / DeepSeek-Coder:} Code-specialized continued pre-training from a general base. Infill (fill-in-the-middle) pre-training objective for code completion.
  \item \textbf{ChatLAW (legal):} Legal corpus with case law, statutes, and regulatory filings.
\end{itemize}

\section{Multimodal Models}

\begin{itemize}
  \item \textbf{Vision-Language Models (VLM):} Connect a vision encoder (CLIP, SigLIP) to an LLM via a projection layer or cross-attention. LLaVA, InternVL, Qwen3-VL.
  \item \textbf{Early fusion (Llama 4):} Text and image tokens processed jointly from the first layer. More parameter-efficient at scale than late-fusion adapters.
  \item \textbf{Vision-Language MoE (Qwen3-VL):} Cross-modal compute allocation---vision and language tokens routed to specialized experts.
  \item \textbf{Omni-modal:} Audio, video, and document understanding in a unified model (Gemini 2.5 Pro, GPT-4o).
\end{itemize}

\section{Agentic AI}

LLMs as agents: perceive environment, plan actions, use tools, and produce observable effects. Training considerations:
\begin{itemize}
  \item End-to-end agent evaluation (SWE-bench, WebArena, GAIA) requires operating in real environments, not just text.
  \item Tool-use fine-tuning: function calling, structured JSON output, code execution.
  \item Multi-agent coordination: multiple specialized LLMs operating in concert.
\end{itemize}
