\chapter{The Future of LLM Training}

\section{Dominant Trends}

\begin{itemize}
  \item \textbf{Post-training dominance:} A strong base model plus well-designed post-training can match models 10$\times$ larger with less careful alignment.
  \item \textbf{Reasoning as a trainable skill:} RLVR + GRPO enable systematic reasoning improvements beyond what SFT achieves.
  \item \textbf{Synthetic data at scale:} Frontier models generate training data for the next generation. Data flywheel dynamics favor incumbents with strong base models.
  \item \textbf{Efficiency breakthroughs:} MoE, FP8, speculative decoding, and quantization collectively enable 10--50$\times$ inference cost reduction vs.\ 2023 baselines.
  \item \textbf{Open-source convergence:} Open models (Llama, Qwen, Mistral) approach closed model performance within 6--12 months of release.
  \item \textbf{Test-time compute scaling:} A new scaling axis complementing parameter and data scaling, trading inference cost for accuracy.
\end{itemize}

\section{Open Challenges}

\begin{itemize}
  \item \textbf{Data scarcity:} Web text quality has plateaued; synthetic data introduces distribution shift and mode collapse risks.
  \item \textbf{Evaluation gaps:} No reliable benchmark for AGI-level tasks, agentic behavior, or long-horizon planning.
  \item \textbf{Safety at scale:} Alignment techniques that work at 7B may not generalize to 1T+ parameter models.
  \item \textbf{Interpretability:} We cannot reliably predict emergent capabilities or explain failure modes.
  \item \textbf{Energy costs:} Frontier training runs consume megawatt-hours; inference at scale requires dedicated data center planning.
  \item \textbf{Legal and regulatory:} EU AI Act, copyright litigation, and data provenance requirements create growing compliance burdens.
  \item \textbf{Multimodal grounding:} Models that understand the physical world through vision, audio, and action remain an open research frontier.
\end{itemize}

\section{The Next Frontier: 2026--2028}

Likely near-term directions: hardware-software co-design (custom AI chips with in-memory compute), world models for physical reasoning, neurosymbolic integration for reliable formal reasoning, and on-device continual learning for personalization without data privacy risks.


% ──────────────────────────────────────────────────────────────────
